Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-10-24 other/confnoncoherent/paper.tex
added to paper.  Filled in first part of the proposition.  It appears difficult to do the second part.  Barry might have some good ideas about this.

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  
A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]  
If $X$ is circularly symmetric, then for any real number $\phi$, the distribution of $X$ is the same as that of $e^{j\phi}X$.  %If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
% \begin{align*}
% \expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
% &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
% &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
% \end{align*}

% We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
% \[
% R e^{j\Phi} = 1 + X.
% \]
% The joint distribution of $R$ and $\Phi$ can be shown to be
% \[
% f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
% \]

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $w_1$ having finite variance and continuous pdf.  Let $y_1,\dots, y_L$ be given by~\eqref{eq:sigmod} and let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$.  Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i},
\end{equation}
and define the continuous function
\[
G(x) = \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\] 
If $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}} \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$.  The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
G(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $G(x)$ is uniquely maximised at $x=0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}

@@

The problem of estimating carrier phase and amplitude from the observation of $L$ noisy $M$-PSK symbols has undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  The least squares estimator can be computed in $O(L\log L)$ operations using the algorithm of Mackenthun~\cite{Mackenthun1994}, and this is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.  The algorithm was later rediscovered by Sweldens~\cite{Sweldens2001}.  In this paper we described the statistical properties of the least squares estimator. We showed that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.

>>>>

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  
A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is $f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z)$.
If $X$ is circularly symmetric, then for any real number $\phi$, the distribution of $X$ is the same as that of $e^{j\phi}X$.  %If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
% \begin{align*}
% \expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
% &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
% &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
% \end{align*}

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
\]

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $w_1$ having finite variance and continuous pdf.  Let $y_1,\dots, y_L$ be given by~\eqref{eq:sigmod} and let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$.  Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
%\begin{equation}\label{eq:RiandPhii}
\[
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i},
\]
%\end{equation}
and define the continuous function
\[
G(x) = \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\] 
If $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}} \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  %This assumption asserts that $G(x) \leq G(0)$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $G(x_i) \rightarrow G(0)$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  
Although we will not prove it here, this assumption is not only sufficient, but also necessary, for if $G(x)$ is uniquely maximised at some $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $G(x)$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  
%The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

% \begin{proposition}\label{prop:contgg2}
% Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having continuous and differentiable pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
% \[
% G(x) =  \expect R \cos\sfracpart{x + \Phi}.
% \]
% Then $G(x)$ is uniquely maximised at $x=0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
% \end{proposition}

@@

The problem of estimating carrier phase and amplitude from the observation of $L$ noisy $M$-PSK symbols has undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  The least squares estimator can be computed in $O(L\log L)$ operations using the algorithm of Mackenthun~\cite{Mackenthun1994}, and this is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.  The algorithm was later rediscovered by Sweldens~\cite{Sweldens2001}.  In this paper we described the statistical properties of the least squares estimator. The phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.

### 2012-10-24 paper.tex
added to paper.  Filled in first part of the proposition.  It appears difficult to do the second part.  Barry might have some good ideas about this.

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{W}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{W}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
\begin{align*}
 \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
 &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
 &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
 \end{align*}
If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
\[
f_{X,Y}(x,y) = \frac{f_Z(\sqrt{x^2 + y^2})}{2\pi \sqrt{x^2 + y^2}}.
\]

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  This assumption asserts that $G(x) \leq G(0)$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $G(x_i) \rightarrow G(0)$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  Although we will not prove it here, this assumption is not only sufficient, but also necessary, for if $G(x)$ is uniquely maximised at some $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $G(x)$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  The proof is given in Appendix~\ref{appendix:prop2proof}.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having continuous and differentiable joint pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $h_1(x)$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $h_2(x)$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}

@@

\subsection{Proof of Proposition~\ref{prop:contgg2}} \label{appendix:prop2proof}

>>>>

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{W}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{W}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
\begin{align*}
 \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
 &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
 &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
 \end{align*}
If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
\[
f_{X,Y}(x,y) = \frac{f_Z(\sqrt{x^2 + y^2})}{2\pi \sqrt{x^2 + y^2}},
\]
from which follows
\begin{equation}\label{eq:fzandfZYwithz}
f_Z(z) = 2\pi z f_{X,Y}(z,0) = 2\pi z f_{X,Y}(0,z)
\end{equation}
for all $z > 0$.

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  This assumption asserts that $G(x) \leq G(0)$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $G(x_i) \rightarrow G(0)$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  Although we will not prove it here, this assumption is not only sufficient, but also necessary, for if $G(x)$ is uniquely maximised at some $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $G(x)$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  %The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  The proof is given in Appendix~\ref{appendix:prop2proof}.

% \begin{proposition}\label{prop:contgg2}
% Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having continuous and differentiable joint pdf $f_{X,Y}(x,y)$ that is decreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
% \[
% h_1(x) = \expect R \cos(x + \Phi) \;\;\; \text{and} \;\;\; h_2(x) =  \expect R \cos\sfracpart{x + \Phi}.
% \]
% Then $h_1(x)$ is uniquely maximised at $x=0$ over the interval $[-\pi,\pi]$ and $h_2(x)$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
% \end{proposition}

@@

% \subsection{Proof of Proposition~\ref{prop:contgg2}} \label{appendix:prop2proof}

% \begin{IEEEproof}
% We first prove the result for $h_1$.  Write
% \begin{align*}
% h_1(x) &= \int_{-\pi}^{\pi} \int_{0}^\infty r \cos(\phi + x) f(r,\phi) dr d\phi \\
% &= \int_{-\pi}^{\pi}  \cos(\phi + x) g(\phi) d\phi,
% \end{align*}
% where $f(r,\phi)$ is the joint pdf of $R$ and $\Phi$, and 
% \[
% g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
% \]
% Since the pdf of $X$ is circularly symmetric, $f(r,\phi)$ and $g(\phi)$ are even in $\phi$ on $[-\pi,\pi]$ (see the discussion after~\eqref{eq:pdfRPhi}).  It is also the case that $g(\phi)$ is unimodal with mode at zero, as a result of Lemma~\ref{lem:gunimodmodzero}.  Write $\cos(\phi + x)$ as a sum of even and odd parts,
% \[
% \cos(\phi + x) = \cos(\phi)\cos(x) - \sin(\phi)\sin(x).
% \]
% Now
% \[
% h_1(x) = \cos(x) \int_{-\pi}^{\pi} \cos(\phi) g(\phi) d\phi
% \]
% since $\int_{-\pi}^{\pi} \sin(\phi) g(\phi) d\phi = 0$ because $\sin(\phi)g(\phi)$ is odd.  Now 
% \[
% K = \int_{-\pi}^{0} \cos(\phi) g(\phi) d\phi > 0
% \] 
% because $g(\phi) > 0$ is increasing on $(-\pi,0)$ and $\cos(\phi)$ is odd and increasing on $(-\pi,0)$.  Thus $h_1(x) = 2K \cos(x)$ is uniquely maximised at zero because $\cos(x)$ is uniquely maximised at zero.
% \end{IEEEproof}

% \begin{lemma}\label{lem:gunimodmodzero}
% Under the assumptions in Proposition~\ref{prop:contgg2} let $f_Z$ be the pdf of $\abs{W}$.  Then $z^{-1} f_Z(z)$ is differentiable and decreasing for $z > 0$ and the function $g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr$ is unimodal on $[-\pi,\pi]$ with mode at $\phi = 0$.
% \end{lemma}
% \begin{IEEEproof}
% From~\eqref{eq:fzandfZYwithz} it follows that $2\pi f_{X,Y}(z,0) = z^{-1} f_Z(z)$ for all $z>0$, and since $f_{X,Y}(z,0)$ is differentiable and nonincreasing in $z$ (by assumption in the statement of Proposition~\ref{prop:contgg2}) we have that $z^{-1} f_Z(z)$ is differentiable and decreasing for $z > 0$.  So the derivative $\frac{d}{dz} \big( z^{-1} f_Z(z) \big) > 0$ for all $z > 0$.   Using~\eqref{eq:pdfRPhi},
% \[
% g(\phi) = \int_{0}^\infty r^2 \frac{f_Z(z)}{2\pi z } dr
% \]
% where $z = \sqrt{r^2 - 2r\cos\phi + 1}$.  Differentiating with respect to $\phi$,
% \[
% g^\prime(\phi) = \sin(\phi) \int_{0}^\infty \frac{r^3}{\pi} \frac{d}{dz}\left( \frac{f_Z(z)}{z} \right) dr
% \]
% and since the term inside the integral is negative, the integral is negative.  Since $\sin(\phi)$ is negative on $(-\pi,0)$ and positive on $(0,\pi)$ it follows that the derivative $g^\prime(\phi)$ is positive on $(-\pi,0)$ and negative on $(0,\pi)$.  Thus $g(\phi)$ is unimodal with mode at $\phi = 0$.
% \end{IEEEproof}
